Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-15 opencl/papers/intro.tex
edits

We provide (via Fortran interfaces in the ForOpenCL library) a mechanism to
call the C OpenCL runtime and enable Fortran programmers to access OpenCL
kernels.  Transformations are supplied that provide a mechanism for converting
Fortran procedures written in the Fortran subset described in this paper to
OpenCL kernels.  We use the ROSE compiler infrastructure \cite{} to develop
these transformations.  ROSE uses the Open Fortran Parser \cite{} to parse
Fortran 2008 syntax and can output (unparse) to OpenCL.  Since ROSE's
intermediate representation (IR) was constructed to represent multiple
languages, it is relatively straightforward to transform high-level Fortran IR
nodes to OpenCL nodes.  Furthermore, the Fortran array syntax maps directly to
one, two, and three-dimensional thread groups in OpenCL.

@@

%% NEW - CER
History has also shown that an investment in rewriting code does not guarantee
success either, as seen in an effort at LANL to modernize a legacy Fortran
code with the the newer C++ POOMA framework.  This massive overhaul effort led
to a code that was both slower and less flexible than the original Fortran
\cite{}.  Similar experiences have occurred in the past, notably during the
devlopment of the Sisal language.

>>>>

We provide (via Fortran interfaces in the ForOpenCL library) a mechanism to
call the C OpenCL runtime and enable Fortran programmers to access OpenCL
kernels.  Transformations are supplied that provide a mechanism for converting
Fortran procedures written in the Fortran subset described in this paper to
OpenCL kernels.  We use the ROSE compiler infrastructure \cite{rose-web} to develop
these transformations.  ROSE uses the Open Fortran Parser \cite{ofp-web} to parse
Fortran 2008 syntax and can output (unparse) to OpenCL.  Since ROSE's
intermediate representation (IR) was constructed to represent multiple
languages, it is relatively straightforward to transform high-level Fortran IR
nodes to OpenCL nodes.  Furthermore, the Fortran array syntax maps directly to
one, two, and three-dimensional thread groups in OpenCL.

@@

%% NEW - CER
History has also shown that an investment in rewriting code does not guarantee
success either, as seen in an effort at LANL to modernize a legacy Fortran
code with the the newer C++ POOMA framework.  This massive overhaul effort led
to a code that was both slower and less flexible than the original Fortran
\cite{pooma-cite}.  Similar experiences have occurred in the past, notably during the
devlopment of the Sisal language.

### 2010-12-15 opencl/papers/programming-model.tex
edits

The static analysis and source-to-source transformations used are very
basic but require the programmer to use a subset of Fortran
that employs a data-parallel programming model.  In particular, it
encourages use of language features that were introduced and
standardized in the Fortran 90 language specification.  In this
section we describe the set of Fortran 90 features that our analysis
and transformation method are based on.

This paper uses four simple features of Fortran that form a language subset
that does not require any language extensions and can be easily transformed
to a lower-level implementation in either CUDA or OpenCL.

@@

Fortran 90 introduced a rich array syntax that allows programmers to
write code that is in terms of whole arrays or subarrays, with data
parallel operators to compute on the arrays.  This has the benefit of
avoiding explicit looping in the code and maintaining a high-level
style that is closer to the original mathematical specification of the
problem being solved.  More importantly from a compilation
perspective, this defers decisions about how to implement these
whole-array operations to a compilation tool.  When faced with a novel
architecture such as modern GPUs that are ideally suited to data parallel
programming models, the fit between Fortran arrays and these systems is
quite clean.

%% NEW - CER
Arrays are first class citizens in Fortran as they are part of the language
type system rather than just a pointer to raw memory as in C. A Fortran array
contains metadata associated with it that describes the array's shape and
size. This metadata is lacking in other languages, such as C, C++, and Java.
Even in the case where multidimensional array data structures are available
(such as in C++ or Java), they exist outside the language standard, limiting
the ability of the compiler to analyze them fully. The existence of this
metadata allows the use of array notation whereby explicit indexing of arrays
is not required. For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays
of the same rank and shape and {\tt s} is a scalar, then the statement

@@

results in the sum of elements of {\tt A} and {\tt s} times the elements of
{\tt B} being stored in the corresponding elements of {\tt C}. The first
element of {\tt C} will contain the value of the first element of {\tt A}
added to the first element of {\tt c*B}.  Note that no explicit iteration over
array indices is needed and that the individual operators, plus, times, and
assignment are applied by the compiler to individual elements of the arrays
independently.  Thus the compiler is able to spread the computation in the
example across any hardware threads under its control.

@@

%% NEW - CER
An example of shown above using array syntax could be refactored to use
the elemental function,

@@

Note that while {\tt foo} is defined in terms of purely scalar
quantities, it can be \emph{applied} to arrays as shown.

While this may seem like a trivial example, such simple functions may be
composed with other elemental functions to perform powerful computations,
especially when applied to arrays.  ForOpenCL transforms elemental functions to inline OpenCL
functions.  Thus there is no penalty for usage of elemental functions and provide
a convient mechanism to express algorithms in simpler segments.

Also note that the plus and times operators shown in the data-statement
example can be seen as an elemental functions, as plus and times are defined
in terms of scalars but can be applied to whole arrays and can return an array
result.

@@

Pure procedures, like elemental functions, must be free of side effects.
(athough they may change the value of array elements in an array argument,
unlike elemental arguments which must have the intent(in) attribute).
The absence of side effects removes ordering constraints from the compiler
allowing it to invoke pure functions out of order.  Procedures and functions
of this sort are also common in pure functional languages like Haskell,
and are exploited by compilers in order to emit parallel code automatically
due to their suitability for compiler-level analysis.

Since pure procedures don't have side effects they a candidates for running on
accelerators in OpenCL.  Currently ForOpenCL transforms pure procedures to
OpenCL kernels that \emph{do not} call other procedures, except for elemental
functions.

@@

Unlike OpenMP, we require limited use of compiler directives, although
two are used to enforce the semantics required by the programming
model.  These directives are:
\begin{itemize}
   \item {\tt \!\$OFP CONTIGUOUS}: specifies that an dummy array variable is
   contiguous in memory. CONTIGUOUS is an attribute in Fortran 2008.
   \item {\tt \!\$OFP KERNEL}: specifies that a pure subroutine can be
     transformed to an OpenCL kernel.
\end{itemize}

Three new procedures (in addition to the intrinsic shift function) are
defined in Fortran that are used in array-syntax operations.  Each
procedure takes a integer array halo argument that specifies the
number of ghost cells on either side of a region, for each dimension.
For example {\tt halo = [left, right, down, up]} specifies a halo for
a two-dimensional region.  These functions are:
\begin{itemize}
   \item {\tt transfer\_halo(array, halo)}: an impure subroutine that
     exchanges halo regions between nodes using MPI or Fortran
     coarrays. Array is the memory stored on the node representing the
     node-local portion of a virtual global array. Not used in this work.
   \item {\tt interior(array, halo)}: a pure function that returns a copy of
     the interior portion of the array specified by halo. Array is
     local to a node.
   \item {\tt region(array, halo)}: a pure function that returns a copy of
     the portion of the array specified by halo.  Array is local to a node.
\end{itemize}

@@

\begin{itemize}
	\item There are no loops or index variables to keep track of.  Off by one
index errors are a common programming mistake.
        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
	\item Semantically the intrinsic functions return arrays by value.  This is usually what the algorithm requires.
	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
\end{itemize}

@@

%% NEW - CER
This example is a solution to Maxwell's equations for the $z$ component of the
magnetic field using Fortran array syntax.  Note that there are no explicit
loops in this example.  The operators {\tt +}, {\tt -}, and {\tt *} are
applied to all of the elements of the three-dimensional arrays {\tt Bz}, {\tt
  Ex}, and {\tt Ey}, individually.  This is why data parallelism has been
called collection-oriented programming by
Blelloch~\cite{blelloch90,rajopadhyedidlacs}.  As the {\tt cshift} function
and the array-valued expressions all semantically returns a value, this style
of programming is also similar to functional programming (or value-oriented
programming). %%~\cite{simonpeytonjones}).  The heart of the solution to
Maxwell's equations is the statement shown in Listing~\ref{lst:dpexample} and
five similar, simple equations.

>>>>

The static analysis and source-to-source transformations used are very
basic and only require the programmer to use a subset of Fortran that
employs a data-parallel programming model.  In particular, it
encourages use of language features that were introduced and
standardized in the Fortran 90 language specification.  In this
section we describe the set of four Fortran 90 features that our analysis
and transformation method are based on.  From these language constructs,
we are able to easily transform them to a lower-level CUDA or OpenCL
implementation.

%This paper uses four simple features of Fortran that form a language subset
%that does not require any language extensions and can be easily transformed
%to a lower-level implementation in either CUDA or OpenCL.

@@

Fortran 90 introduced a rich array syntax that allows programmers to
write statements that are in terms of whole arrays or subarrays, with
data parallel operators to compute on the arrays.  This has the
benefit of avoiding explicit looping in the code and maintaining a
high-level style that is closer to the original mathematical
specification of the problem being solved.  More importantly from a
compilation perspective, this defers decisions about how to implement
these whole-array operations to a compilation tool.  Sophisticated
loop analysis to identify parallelism within the loop are not
necessary.  When faced with a novel architecture such as modern GPUs
that are ideally suited to data parallel programming models, the fit
between Fortran arrays and these systems is quite clean.

%% NEW - CER
Arrays are first class citizens in Fortran instead of a basic pointer
to raw memory as in many other compiled languages.  A Fortran array
contains metadata associated with it that describes the array's shape
and size. It should be noted that other languages that provide rich
array data types and whole array operations may also be suitable
targets for the transformations described in this paper.  
In addition to metadata about the structure of the arrays, operators
are provided that allow programmers to write expressions based on the
whole array instead of explicit indexing and application of operations
to individual elements.  For example, if {\tt A}, {\tt B}, and {\tt C}
are all arrays of the same rank and shape and {\tt s} is a scalar,
then the statement

@@

results in the element-wise sum of {\tt A} and {\tt s} times the elements of
{\tt B} being stored in the corresponding elements of {\tt C}. The first
element of {\tt C} will contain the value of the first element of {\tt A}
added to the first element of {\tt c*B}.  Note that no explicit iteration over
array indices is needed and that the individual operators, plus, times, and
assignment are applied by the compiler to individual elements of the arrays
independently.  Thus the compiler is able to spread the computation in the
example across any hardware threads under its control.

@@

%% NEW - CER
For example, the basic array operation shown above could be refactored into
an elemental function,

@@

Note that while {\tt foo} is defined in terms of purely scalar
quantities, it can be \emph{applied} to arrays as shown.  While this
may seem like a trivial example, such simple functions may be composed
with other elemental functions to perform powerful computations,
especially when applied to arrays.  Our prototype tool transforms
elemental functions to inline OpenCL functions.  Thus there is no
penalty for usage of elemental functions and provide a convient
mechanism to express algorithms in simpler segments.

%Also note that the plus and times operators shown in the data-statement
%example can be seen as an elemental functions, as plus and times are defined
%in terms of scalars but can be applied to whole arrays and can return an array
%result.

@@

Pure procedures, like elemental functions, must be free of side
effects.  Unlike elemental functions that require arguments to have an
{\tt intent(in)} attribute, they may change the contents of array
arguments that are passed to them.  The absence of side effects
removes ordering constraints that could restrict the freedom of the
compiler to invoke pure functions out of order and possibly in parallel.
Procedures and functions of this sort are also common in pure
functional languages like Haskell, and are exploited by compilers in
order to emit parallel code automatically due to their suitability for
compiler-level analysis.

Since pure procedures don't have side effects they a candidates for
running on accelerators in OpenCL.  Currently our prototype tool
transforms pure procedures to OpenCL kernels that \emph{do not} call
other procedures, except for elemental functions.

@@

Unlike OpenCL, we require limited use of compiler directives, although
two are used to enforce the semantics required by the programming
model.  These directives are:

\begin{itemize}

\item {\tt \!\$OFP CONTIGUOUS}: specifies that an dummy array variable is
  contiguous in memory. CONTIGUOUS is an attribute in Fortran 2008.

\item {\tt \!\$OFP KERNEL}: specifies that a pure subroutine can be
  transformed to an OpenCL kernel.

\end{itemize}

Three new procedures (in addition to the intrinsic shift function) are
defined in Fortran that are used in array-syntax operations.  Each
procedure takes a integer array halo argument that specifies the
number of ghost cells on either side of a region, for each dimension.
For example {\tt halo = [left, right, down, up]} specifies a halo for
a two-dimensional region.  These functions are:

\begin{itemize}

\item {\tt transfer\_halo(array, halo)}: an impure subroutine that
  exchanges halo regions between nodes using MPI or Fortran
  coarrays. Array is the memory stored on the node representing the
  node-local portion of a virtual global array. Not used in this work.

\item {\tt interior(array, halo)}: a pure function that returns a copy of
  the interior portion of the array specified by halo. Array is
  local to a node.

\item {\tt region(array, halo)}: a pure function that returns a copy of
  the portion of the array specified by halo.  Array is local to a node.

\end{itemize}

@@

\begin{itemize}
\item There are no loops or index variables to keep track of.  Off by
  one index errors and improper handling of array boundaries are a
  common programming mistake.
\item The written code is closer to the algorithm, easier to
  understand, and is usually substantially shorter.
\item Semantically the intrinsic functions return arrays by value.
  This is usually what the algorithm requires.
\item Because pure and elemental function are free from side effects,
  it is easier for a compiler to schedule the work to be done in
  parallel.
\end{itemize}

%% MJS - mention that these features were designed for parallelism in
%% the first place?

@@

%% NEW - CER

This example is a solution to Maxwell's equations for the $z$
component of the magnetic field using Fortran array syntax.  Note that
there are no explicit loops in this example.  The operators {\tt +},
{\tt -}, and {\tt *} are applied to all of the elements of the
three-dimensional arrays {\tt Bz}, {\tt Ex}, and {\tt Ey},
individually.  This is why data parallelism has been called
collection-oriented programming by
Blelloch~\cite{blelloch90,rajopadhyedidlacs}.  As the {\tt cshift}
function and the array-valued expressions all semantically returns a
value, this style of programming is also similar to functional
programming (or value-oriented
programming).

%%~\cite{simonpeytonjones}).  The heart of the solution
%to Maxwell's equations is the statement shown in
%Listing~\ref{lst:dpexample} and five similar, simple equations.
